---
title: "Footfall Project"
description: "The Effect of Covid-19 on Pedestrian Footfall in Dublin City Centre."
author: "Niall Ryan"
  
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
library(dplyr)
library(ggmap)
library(mapview)
library(sf)
library(tidyr)
library(tidyverse)
library(leaflet)
library(widgetframe)
```

## Introduction

This report will explore the effect of Covid-19 on pedestrian footfall count in Dublin City Centre through the use of exploratory data analysis. This analysis will focus on the pedestrian footfall counts of people walking past certain locations in Dublin City Centre in 2020 from January to the end of the Novemeber.

All of the data was extracted from the Smart Dublin website via the Dublin City Council who collected the raw data in a csv file which we loaded in so we could produce an anlysis of this data.


## Importing and Inspecting the Dataset

Before exploratory analysis can begin, the data must be imported and its structure inspected. The data was imported using read.csv and inspected firstly using the str() function. 


```{r}

suppressMessages(library(tidyverse))

d<- read.csv("data/jan-nov-2020-ped-data.csv")
 
str(d)

```



The str() function shows there are 8040 observations across 24 columns, column 1 containing chr values corresponding to date and time, and the remaining 23 columns giving integer values of pedestrain count for specific locations in Dublin.

No values for pedestrian count were obtained for Dawson Street, possibly due to a malfunctioning counter. This column will be removed from further analysis.





```{r}
summary(d) 

```


Next, the summary() function shows that there are many missing/NA values in all columns but a very large portion of NA values were in Capel Street, Mary Street, Dame Street, College Green, Henry Street and Grafton Street columns.These values will be omitted in the analysis also.



## Cleaning the Data


The next task before performing any exploritory data analysis on this dataset was to clean and format the data. The first thing that needed to be done was to remove the problematic Dawson Street column which was entirely made up of missing values. After this, the replacement Dawson Street counter was set as Dawson Street and other columns were renamed in order to make them more presentable in our analysis.



```{r}
ped<-d[,-18] 



ped<- ped %>% rename(`Capel Street`= Capel.Street...Removed.from.site.20.10,`Dame Street`= Dame.Street..Counter.Missing.,`Dawson Street`= Dawson.Street.Replacement)





```


However, there were still missing values present that needed to be omitted from the dataset.The median was imputed to replace the missing values. To ommit them  a function was created to replace the missing value with the median value of each street. We decided to impute the Median for the missing values as it is not going to be skewed by any outliers and in this case due to the type of count values being recorded, it is therefore a more accurate prediction than imputing the values with the mean(which can be skewed by outliers)or with 0 values(doesn't give a true representation of the data).



```{r}
replaceNA <-function(x){
  ifelse(is.na(x),median(x,na.rm =TRUE),x)
}

ped$O.Connell.St.Outside.Pennys <- replaceNA(ped$O.Connell.St.Outside.Pennys)
ped$O.Connell.St.Outside.Clerys<-replaceNA(ped$O.Connell.St.Outside.Clerys)
ped$Mary.Street<-replaceNA(ped$Mary.Street)
ped$`Capel Street`<-replaceNA(ped$`Capel Street`)
ped$Aston.Quay<-replaceNA(ped$O.Connell.St.Outside.Clerys)
ped$Grafton.Street...CompuB<- replaceNA(ped$Grafton.Street...CompuB)
ped$Talbot.Street.North<- replaceNA(ped$Talbot.Street.North)
ped$Doilier.Street..Burgh.Quay<- replaceNA(ped$Doilier.Street..Burgh.Quay)
ped$`Dawson Street`<- replaceNA(ped$`Dawson Street`)
ped$`Dame Street`<- replaceNA(ped$`Dame Street`)
ped$Talbot.Street.South <- replaceNA(ped$Talbot.Street.South)
ped$O.Connell.St..Parnell.St...AIB<- replaceNA(ped$O.Connell.St..Parnell.St...AIB)
ped$Grafton.Street...Nassau.Street...Suffolk.Street<- replaceNA(ped$Grafton.Street...Nassau.Street...Suffolk.Street)
ped$College.Green..Bank.Of.Ireland<- replaceNA(ped$College.Green..Bank.Of.Ireland)
ped$Henry.Street<- replaceNA(ped$Henry.Street)
ped$Westmoreland.Street.East<-replaceNA(ped$Westmoreland.Street.East)
ped$Liffey.Street<- replaceNA(ped$Liffey.Street)
ped$Westmoreland.Street.West<- replaceNA(ped$Westmoreland.Street.West)
ped$Grafton.Street<-replaceNA(ped$Grafton.Street)
ped$Bachelors.Walk<-replaceNA(ped$Grafton.Street)
ped$College.Green...Church.Lane<-replaceNA(ped$College.Green...Church.Lane)
ped$College.Green...Dame.St.Side<-replaceNA(ped$College.Green...Dame.St.Side)




```


After the missing values were ommitted from the dataset, additions could be made and some last formatting issues needed to be dealt with before the exploritory data analysis could occur. A total column was added to the dataset to aid us in graphing the total amount of pedestrian footfall for each hour of every day. Also date and time were seperated and date was given a date format so we could make use of this time series data in our analysis.




```{r}


ped <- mutate(ped, total =O.Connell.St.Outside.Pennys+ O.Connell.St.Outside.Clerys+ ped$Mary.Street+ `Capel Street`+ Aston.Quay+ Grafton.Street...CompuB+ Talbot.Street.North+ Doilier.Street..Burgh.Quay+ `Dawson Street`+ `Dame Street`+ Talbot.Street.South+ O.Connell.St..Parnell.St...AIB+ Grafton.Street...Nassau.Street...Suffolk.Street+College.Green..Bank.Of.Ireland+ Henry.Street+ Westmoreland.Street.East+ Liffey.Street+ Westmoreland.Street.West+Grafton.Street+Bachelors.Walk+ College.Green...Church.Lane+ College.Green...Dame.St.Side )


ped <- separate(ped,Date...Time, c("Date", "Time"), sep=" ", remove=F)

ped$Date<-as.Date(ped$Date)







```








## Overview  
Using the DCC footfall counter locations data provided, the following map of Dublin and the counter locations was produced using a google maps API. Firstly, the Dawson Street location was removed from the data as a Dawson Street replacement counter at a different location was included. Next, a base map of Dublin was formed using the getmap function, and the Latitude and Longitude values from the original dataset was converted to a geom value, "counters" with the st_assf function in the sf library. Qmplot was then used to map the counter locations.
The locations of the counters are clearly clustered around Dublin city center, except for 1 at Talbot Street North which is seperate from the others. As the Talbot Street North counter is further outside of the city center, it is unsurpriseing that it has a relatively low median value for pedestrian count at 187.

```{r}
location<- read.csv("data/dcc-footfall-counter-locations-14082020.csv")
location<-location[-18,]

suppressMessages(library(ggmap))

ggmap::register_google(key = "AIzaSyAbeYjOXHG-rGqRbzQ66JaB9d5iMfXTSxM")
dublin <- get_map(location = "dublin", zoom = 13)

counters <- st_as_sf(location, coords = c("Longitude", "Latitude"), crs = 4326) 

qmplot(Longitude,  Latitude, data =location,  source="google",maptype="roadmap",color = I("red"),zoom = 13)
```
The following heatmap is a repesentation of the median value of pedestrian count on each street, which are numbered 1-22. As shown, Mary Street(3) had the lowest median value of pediestrians at 53, and Henry Street(15) had the highest overall at 759. The remaining streets are also colour graded from Yellow to Red based on their median values.

```{r}
Medians<-c(lapply(ped[4:25], median))
location<- mutate(location, Medians)
Medians<-as.data.frame(Medians)
Medians_t <- t(Medians)
Medians_t<-as.data.frame(Medians_t)

Street.no<-c(1:22)
Street.no<-as.data.frame(Street.no)
Medians_t<-mutate(Medians_t, Street.no)
Medians_t<-as.matrix(Medians_t)
heatmap(Medians_t, scale="none")
```





## Comparing Lockdown footfall 

Focusing inon the data, we wanted to look specifically at the effect of pedestrian numbers during the lockdown imposed at the end of March and the lockdown imposed just in Dubin in September. To make an accurate comparison, we looked at the first full month of each lockdown, filtering our dataset accordingly. 


```{r}

street.nam <-rep(colnames(ped)[4:25])
street.nam <- as.factor(street.nam)

April<- colSums(ped[months(ped$Date)=='April',c(4:25)])

lock1 <- data.frame(street.nam,totapr=c(April))

lock1$Month<- c(rep('April',22))
lock1$Month<-as.factor(lock1$Month)



ggplot(data=lock1, aes(x=street.nam, y=totapr)) +
  geom_col(col='green')+coord_flip()+
  theme_classic()+
  labs(title = "Lockdown - Nationwide Level 5",
          subtitle = "Number of Pedestrians during this lockdown - April",
           y = "Total Pedestrians",
           x = "Street")+
  scale_y_continuous(breaks=c(50000,150000,250000,350000,450000))
          


```


In the graph produced above, we see the pedestrian levels during the first month of the level 5 lockdown. We see the effect on pedestrian on each street showing a stark decrease in volume with 80% of the streets data recorded from this month maintaining a total pedestrian count below 150,000 level. This shows there was a compliance towards the Government's “Stay at home” message. 

 The low numbers could also be down to the fact that all non-essential businesses had to shut down during this period, so there was nothing really to entice people to walk around Dublin City Centre. 

 

In comparison, we looked at the lockdown imposed on Dublin (only) later in the year, which was imposed in September. This was to see had there been any changes in behaviour in the form of pedestrian numbers from the first full month during the lockdown in April, we can see the second lockdown graph shown below.  





```{r}


September<- colSums(ped[months(ped$Date)=='September',c(4:25)])

lock2 <- data.frame(street.nam,totsep=c(September))

lock2$Month<- c(rep('September',22))
lock2$Month<-as.factor(lock2$Month)




ggplot(data=lock2, aes(x=street.nam, y=totsep)) + geom_col(col='yellow')+coord_flip()+theme_classic()+labs(title = "Dublin Only Lockdown -  Level 3", 

           subtitle = "Number of Pedestrians during this lockdown - September", 

           y = "Total Pedestrians", 

           x = "Street") 




```

We see there has been a change in numbers walking around as the total amount has increased for each street from April. However, there is a number of factors in which to explain this increase in numbers: 

-The lockdown level in September was set to Level 3 so people had less restrictions to abide by and therefore had more freedom (Level 5 – stay at home,maximum 5km away  vs Level 3 – stay within your county) 

-Retail shops and restaurants were able to open in level 3 restrictions so people had more reason to go into Dublin City Centre and walk around then they did in April. 




## Comparing post Lockdown Footfall 


After looking at the number of pedestrians during both lockdowns, next we focused our attention on the month post lockdown to see if there there was increase or decrease after both lockdowns were lifted and then compare these months after in this case May and October to see if there was a change in pedestrian volume between them.


```{r echo=FALSE, fig.width=14, fig.height=9,out.width="1026",out.height="528"}

street.names <-rep(colnames(ped)[4:25],2)
street.names <- as.factor(street.names)

MaySums <- colSums(ped[months(ped$Date)=='May',c(4:25)])
OctoberSums <- colSums(ped[months(ped$Date)=='October',c(4:25)])

compare <- data.frame(street.names,totsums=c(MaySums,OctoberSums))

compare$Month <- c(rep('May',22),rep('October',22))
compare$Month <-as.factor(compare$Month)



ggplot(compare,aes(x='',y=totsums,fill=street.names))+
  geom_bar(stat='identity',position='dodge',col='white')+
  facet_grid(.~Month)+
  ylab('Number of pedestrians')+
  scale_y_continuous(breaks=c(100000,200000,300000,400000,500000,600000,700000,800000,900000,1000000,1100000))+
  theme_classic()+
  theme(legend.text = element_text(size =8,face='bold'),legend.position='bottom')+
  ggtitle('Number of pedestrains in May and Octorber(after both Lockdowns)')+
  xlab('Month')+
  guides(fill = guide_legend(title = NULL))



```



The graph above shows the difference between the number of pedestrians in the first month post lockdowns. We see in May that people were still reluctant to go out into Dublin continuing the same low numbers trend we saw during the lockdown in April, which we seen in the previous section of the report, even after the lockdown was lifted. 

Our assumptions would suggest because there was an increase in the number of pedestrians between the first lockdown and the second lockdown during these restrictions, we would assume to see that we would see the same trend post lockdown, and so there was no surprise that there was an increase between post lockdowns when restrictions were lifted as October shows an increase number of pedestrians compared to May, with numbers doubling in Henry street, Grafton Street and Talbot Street South when comparing the two months. 


```{r}

sumped<- ped[months(ped$Date)=='May'|months(ped$Date)=='October',c(1,2,3,26)]

 


sumped$month <- c(rep('May',744),rep('October',744))

 

sumped %>% group_by(Date,month) %>%
summarise(pedmean = mean(total)) %>%
ggplot(aes(x = Date, y=pedmean,color=month))+
  geom_point(size=2,color='pink')+
  geom_line(aes(group= factor(month)))+
  ggtitle('The Average Pedestrains Post Lockdown - Trends')+
  ylab('Average footfall')+theme_classic()




```



The line graph above shows the average amount of pedestrians in May and October which show the first month after two different levels of lockdowns. The latter lockdown was less restrictive than the former, so it could be predicted that the mean value in 31 days after  the first lockdown did fluctuate slightly, but less intense than October.  

After the first lockdown the pedestrian volume in Dublin City on average remained between about 2800 and 5500 people per day. However, after the second lockdown, the average amount raised to almost twice that number of pedestrians in May. 
We see that the average amount of pedestrians decreased dramatically after the 21st of October, the main reason for this being Ireland once again went into a level 5 nationwide lockdown.  



## Conclusion

This report looked to answer our research question which was what effect did Covid-19 have on pedestrian footfall in Dublin City Centre in 2020. What we learned from carrying out this exploratory data analysis is that when Covid-19 struck it had a massive effect on the numbers of pedestrians as this dramatically decreased when the first lockdown in at the end of March start of April was imposed and even when lockdown restrictions were lifted the numbers remained low showing there was compliance about decreasing the spread of the virus.  

In the latter half of the year, we see a steady increase in pedestrian numbers even when Dublin was placed into a level 3 lockdown. When this lockdown was lifted numbers continued to increase, we assume that this was mainly due to non-essential business opening up and so people were enticed into going into Dublin City Centre, this increase continued until the end of October when there was a decrease as a nationwide level 5 lockdown was introduced.  